---
title: "Characteristics of STEP for R36"
author: "Andrea Molino"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: cosmo
    toc: yes
  pdf_document:
    toc: yes
---

This first section calculates the overall STEP participant characteristics for my R36 resubmission. 

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(data.table)
library(table1)

n_due     <- 12928
n_overdue <- 8279
n_unknown <- 9942
n_total   <- n_due + n_overdue + n_unknown

prop_due     <- n_due / n_total
prop_overdue <- n_overdue / n_total
prop_unknown <- n_unknown / n_total
prop_total   <- prop_due + prop_overdue + prop_unknown    # just did this to verify, must be 1 

print(prop_due*100)
print(prop_overdue*100)
print(prop_unknown*100)


```


```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### Age
# Although mean age is calculated in the STEP paper, it's stratified by CC history status, and want 1 unifying age mean 
ageweigh_due     <- prop_due * 46.5 
ageweigh_overdue <- prop_overdue * 47.4
ageweigh_unknown <- prop_unknown * 43.8
age_mean         <- ageweigh_due + ageweigh_overdue + ageweigh_unknown

print(age_mean)

```

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### Race
race_total <- 12307 + 7104 + 5959

prop_asian     <- (1521 + 816 + 1067)/race_total
prop_black     <- (600 + 370 + 442)/race_total
prop_morethan1 <- (453 + 278 + 183)/race_total
prop_white     <- (9039 + 5230 + 3828)/race_total

prop_asian*100
prop_black*100
prop_morethan1*100
prop_white*100

# WA proportions 2022: https://ofm.wa.gov/washington-data-research/statewide-data/washington-trends/population-changes/population-race
prop_asian_WA     <- 808748/7864400
prop_black_WA     <- 352936/7864400
prop_morethan1_WA <- 786160/7864400
prop_white_WA     <- 5658184/786440

prop_asian_WA*100
prop_black_WA*100
prop_morethan1_WA*100
prop_white_WA*100

```


```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### Ethnicity 
ethnic_total <- 11994 + 6473 + 3316
prop_hisp <- (846 + 401 + 458)/ethnic_total
prop_hisp*100

# WA proportions 2022: https://ofm.wa.gov/washington-data-research/statewide-data/washington-trends/population-changes/population-hispaniclatino-origin
prop_hisp_WA <- 0.145


```


```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### BMI 
# Percentage with BMI 30 or above
bmi_total  <- 12298 + 5527 + 4256
prop_obese <- (2259 + 1131 + 828 + 1269 + 716 + 530+ 1276 + 800 + 588)/bmi_total
prop_obese*100

```

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### Comorbidity index 
# Percentage with 1 or more on the comorbidity scale 
comorb_total   <- 10752 + 7269 + 9155 + 1925 + 852 + 686 + 157 + 103 + 63 + 94 + 55 + 38
prop_comorbgt1 <- (1925 + 852 + 686 + 157 + 103 + 63 + 94 + 55 + 38)/comorb_total
prop_comorbgt1*100


```


```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### Travel time 
# Percentage that travel 30 minutes or more to clinic 
travel_total <- 12924 + 8277 + 9940
prop_travelgt30 <- (1989 + 1297 + 1286)/travel_total
prop_travelgt30*100


```


```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### SVI 
# Percentage in the most vulerable population 
SVI_total    <- 4185 + 2467 + 2642 + 3862 + 2550 + 2903 + 3133 + 2036 + 2536 + 1748 + 1226 + 1861 
prop_SVIhigh <- (1748 + 1226 + 1861)/SVI_total
prop_SVIhigh*100


```


This second section estimates the number of participants in each of the analyses for Aim 2, specifically HbA1c, mammogram, and colorectal cancer screening, as these will be sub-populations as not everyon needs testing for diabetes or is eligible for the other cancer screenings based on age. These values are estimated based on the Kariya paper, taking the percentage in each analysis from that analysis and applying it to the total number of participants in the STEP study who were CC screening adherent (i.e., my Aim 2 population). 


```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### HbA1C
n_home     <- 14218

n_hba1c_aim2 <- (1036/n_home)*n_due
n_mamm_aim2  <- (3500/n_home)*n_due
n_crc_aim2   <- (4482/n_home)*n_due

print(n_hba1c_aim2)
print(n_mamm_aim2)
print(n_crc_aim2)


```








